Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-07-01 Paper/paper.tex
s6 prelim done, need new figs

\red{SAY HERE THAT WE ONLY USE THE $l=|m|=2$ modes.}
% 
\subsection{Waveform Models}\label{s2:waveforms}

@@

% \textbf{Multiple identical sources at low SNR: }
\label{s2:identical_multiple}

@@

While we discussed Eq.~\ref{eq:rmsSigmaIdenticalSources} for a population
of optimally oriented sources, it is valid more generally if the population is
distributed uniformly in {\it effective} volume~\footnote{{\it effective} 
distance $D_\mathrm{eff}$ 
is a combination of distance to the source, its orientation, and its sky
location angles, and has a one-to-one correspondence with SNR for non-precessing
systems, for which these angles are constant in time. Effective volume is 
proportional to the $D_\mathrm{eff}^3$}~\cite{Markakis:2010mp}. However, it 
still only applies to sources with identical masses and spins, and we 
overcome this limitation by performing a fully Bayesian analysis next.

% There are two main caveats to this rudimentary calculation, first outlined in 
% Ref.~\cite{Markakis:2010mp}: (i) it applies to identical sources,
% which is astrophysically next to impossible to achieve exactly, but could perhaps happen 
% approximately; (ii) the combined average measurement error $\sigma_{avg}$ has been defined
% as in the first equality in Eq.~\ref{eq:rmsSigmaIdenticalSources}. To overcome both,
% we next perform a Bayesian population analysis.
%

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
% 
Imagine that we have $N$ stretches of data, $d_1, d_2, \cdots, d_N$, each 
containing a single signal emitted by an NSBH binary. Each of these signals can
be characterized by the following binary parameters:
$\,\vec{\theta} = \{\mbh, \mns, \chibh, \chins, \rho\}\cup\{\lambdans\}$. Note
that we have folded in the luminosity distance, inclination angle of the 
binary's orbital angular momentum with respect to the plane of the detector,
and source's sky location angles into a single parameter $\rho$, which is the 
SNR of the received signal. This is possible because of our simplification of 
using only the dominant $l=|m|=2$ modes of the GW strain~\footnote{The 
$l=|m|=2$ multipoles of the GW strain, when decomposed in a spin $-2$ weighted
spherical harmonic basis, contain more than $99\%$ of the total signal power}
that makes various angle parameters completely degenerate with distance. 
Further, let $K$ denote all of our collective prior knowledge, except for 
expectations on binary parameters themselves, which enter the following 
calculation explicitly. For instance, $K$ includes our assumption that all
NS's in a single population have the same deformability parameter $\lambdans$, 
and its cumulative measurement is therefore possible.
% 
Using Bayes' theorem, the measured probability distribution function for 
$\lambdans$ from $N$ observations is given by~\footnote{%
Eq.~\ref{eq:lambdaMultiple} includes the following implicit factor to normalize
its right-hand side: $\dfrac{\prod_i p(d_i|K)}{\int p(\lambdans)\,p(d_1, d_2, \cdots, d_N | \lambdans; K)\,\D\lambdans}$}
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{1-N}\prod_{i=1}^N p(\lambdans | d_i, K),
\end{equation}
where $p(\lambdans|K)$ is the prior expectation for $\lambdans$, and all $N$ 
observations are taken to be mutually independent. A priori, we assume
that no particular value of $\lambdans$ is preferred over another, and 
$\lambdans\leq 4000$, i.e.
\begin{equation}\label{eq:lprior}
 p(\lambdans | K) = \dfrac{1}{4000}\,\mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right).
\end{equation}
% 
In the second set of terms in Eq.~\ref{eq:lambdaMultiple} (of the form 
$p(\lambdans | d_i, K)$), each is the probability distribution for $\lambdans$
inferred {\it a posteriori} from the \textit{i}-th observation in isolation. 
They are calculated by integrating
\begin{equation}\label{eq:margpost}
 p(\lambdans | d_i, K) = \int \D \vec{\theta}\, p(\vec{\theta}, \lambdans | d_i, K),
\end{equation}
where $p(\vec{\theta}, \lambdans | d_n, K)$ is the inferred joint probability 
distribution of all source parameters $\vec{\theta}\cup\{\lambdans\}$ for the 
$i$-th event, as given by Eq.~\ref{eq:postprob}. By substituting
Eq.~\ref{eq:lprior}-\ref{eq:margpost} into Eq.~\ref{eq:lambdaMultiple}, we
calculate the probability distribution for $\lambdans$ as measured using $N$
independent events.

Our goal is to understand the improvement in our measurement of $\lambdans$
with the number of recorded events. To do so, we simulate populations of $N$
observations each, and quantify what we learn from each successive observation 
using Eq.~\ref{eq:lambdaMultiple}. For each population, we determine how
rapidly our median estimate for $\lambdans$ converges to the true value,
and how rapidly our confidence intervals for the same shrink, with increasing
$N$. In order to generate each population, we first fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ fixed value chosen from
$\{500,800,1000,1500,2000\}$. Each event is generated independently, sampling
the remaining source parameters uniformly from the following ranges: (i)
mass-ratio $q\in[2,5]$, (ii) BH spin $\chibh\in[0, 0.75]$, (iii) effective 
distance from within a spherical volume whose radius is given by the lowest 
detectable SNR (which we take as $\rho=10$). We remind ourselves that
once binary parameters are fixed, there is a one-to-one mapping between
effective distance and SNR. Additionally, we consider an alternate population
paradigm (paradigm B, with paradigm A being the standard one described 
above), where populations are generated as above but further restricting
mass-ratios such that source BH masses lie {\it outside} the astrophysical 
mass-gap~\cite{Bailyn:1997xt,Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}.
Finally, in order to mitigate computational cost, we make an additional 
simplification. We perform full Bayesian parameter estimation for a set of 
simulated signals whose parameters are the vertices of the following regular
hypercubic grid (henceforth ``G''):
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
The final population is given by substituting all of the $N$ events, generated
as above, with their respective nearest neighbours on the parameter grid G.

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show illustrative results
for a single 
population of $N=80$ events. The neutron star deformability is fixed at
$\lambdans=800$ for all events. In the left panel, each curve shows the 
probability distributions for $\lambdans$ as inferred from $N$ events, with $N$
ranging from $1-80$. We also mark the $90\%$ confidence intervals associated
with each of the probability distribution curves. From the figure, we
can visualize information accumulation. The first few observations do not have
enough information to bound $\lambdans$ (with $90\%$ confidence) much more than
our prior from Eq.~\ref{eq:lprior} does. However, the median of these
cumulative distributions appears to track the true value of $\lambdans$ 
rapidly.
% 
In the right panel, we present information derived from the left panel.
The line-circle curve shows the measured median from $N$ observations, with
the latter on the $x$-axis. The pair of dashed (dotted) horizontal lines mark
$\pm25\%$ ($\pm50\%$) error bars. At each $N$, the range spanned by the 
filled region is the $90\%$ confidence interval deduced from the same 
events. This figure somewhat quantifies the qualitative deductions we made
from the left panel. We find that the measured median does track the true value
quickly, reaching $10\%$ accuracy with $10-15$ observations. With the same
information, our confidence intervals also shrink to $\pm 25\%$.
% 
In Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0} we show further results from $4$
independent populations, with the source $\lambdans$ alone changing between
them. The sampled values, i.e.$\lambdans=\{500,1000,1500,2000\}$, 
range from the least to the most deformable neutron stars. As in the 
right panel of Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}, the line-circle curves
track the median measured $\lambdans$, while the filled regions show
the associated $90\%$ confidence intervals. From the figure, we observe
the following: (i) the shrinkage of confidence interval widths with increasing
$N$ happens in a similar manner for all 
populations, (ii) within $5-10$ observations, median measured $\lambdans$ for
all populations lie within $\pm10\%$ of the true values, and (iii) it takes
approximately $20$ events to distinguish definitively (with $90\%$ confidence)
between deformable NSs with $\lambdans=2000$ and compact NSs with 
$\lambdans=500$, or equivalently to distinguish between hard, moderate and soft
nuclear equations of state. This is comparable to what has been found for
monolithic binaries of neutron stars~\cite{DelPozzo:13,Chatziioannou:2015uea,
Agathos:2015a}.

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization) about the
measurability of $\lambdans$. We mitigate this by drawing multiple
realizations of population generation (i.e. multiple identically constructed
populations). Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} shows the first
results for multiple populations. Lets focus on the left panel first. We
show the median measured value of $\lambdans$ from $100$ independent 
populations with fixed $\lambdans=2000$ as the top set of curves. The one 
highlighted (in green) curve corresponds to the population discussed above.
Similarly, we generate sets of $100$ populations with $\lambdans=500-1500$, and
show the median measured $\lambdans$ as four distinct groups of curves in
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} (with the single highlighted
curve in each group corresponding to
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}).
Dash-dotted horizontal lines demarcate $\pm50\%$ error intervals around
the true $\lambdans$ values. All of these populations are drawn under
paradigm A (i.e. with BH masses allowed to sample the mass-gap that spans
$2-5M_\odot$).
% 
From the figure, we can conclude with a high likelihood that our measured
$\lambdans$ values will be within $10\%$ of the real value in $15-20$ events
for less deformable neutron stars with $\lambdans\leq 1000$, and in as few as
$10$ events for more deformable stars with $\lambdans\geq 1500$.
% 
The right panel of the figure is identical to the left panel, but drawn under
population paradigm B (no BHs alloewd within the mass-gap). Under this
paradigm, we expectedly find that information accumulation is slower. It would
take $20-40$ detections with $\rho\geq10$ under this paradigm, for our (median)
measured $\lambdans$ to lie within $10\%$ of the true value (fewer for 
more deformable NSs, as one would expect).
% 
This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

Further, we investigate the statistical uncertainties associated with
$\lambdans$ measurements. We use $90\%$ confidence intervals as our measure of
the same. The left column of 
Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne} shows results
under paradigm A, where BH mass priors include the mass-gap. In each panel
we show the width of the $90\%$ confidence interval measured from $N$ events
(with $N$ on the $x$-axis), normalized by the true value of $\lambdans$ for
the population. In grey are shown results from each of $100$ population
realizations, with the single population highlighted in color corresponding
to the one we focused on in 
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.
The inset shows the same, except that the ordinate is {\it not} normalized by
the true value of $\lambdans$. From top to bottom, the population $\lambdans$
decreases from $\lambdans=2000\rightarrow 500$, corresponding to decreasingly
deformable NSs with softer equations of state. We observe the following:
(i) for moderately-hard to hard equations of state with $\lambdans\geq 1000$,
we can constrain $\lambdans$ within $\pm 50\%$ using only $10-20$ events, and
within $\pm 25\%$ with $25-40$ events; (ii) for softer equations of state with
$\lambdans<1000$, we will achieve the same accuracy with $20-30$ and $50+$ 
events, respectively; and (iii) for the first $5$ or so observations, our
measurement confidence spans the entire prior allowed range 
$\lambdans\in[0,4000]$, as shown by the plateauing towards the left edge in
each panel's inset.
% 
In the right column, each panel panel is identical to its corresponding left
one, with the difference that populations are drawn under paradigm B, which
does {\it not} allow for BH masses to fall within the mass-gap. We find that
for NSs with $\lambdans\leq 1000$, it would take $25-40$ events to
constrain $\lambdans$ within $\pm 50\%$ and $50+$ events to constrain it
within $\pm 25\%$. This is somewhat slower than paradigm A, as is to be
expected since here we preclude the lowest mass-ratios, which correspond to
signals with largest tidal signatures. For $\lambdans>1000$ we find that we
can constrain $\lambdans$ within $\pm 50\%$ with a similar number of events as
for paradigm A, but will need more ($30-40$, as compared to $25-40$) events
to further constrain it to within $\pm 25\%$ of the true value.
% 
These are some of the primary findings of this paper.

@@

To summarize, in this section we study the improvement in our measurement of 
NS deformability parameter $\lambdans$ with an increasing number of events. We
do so by simulating plausible populations of disrupting NSBH binaries (with
$\rho\geq 10$). We find that:
(i) for more deformable neutron stars (harder equation of states), the median
measured value of $\lambdans$ comes within $10\%$ of the true value with as 
few as $10$ events, while achieving the same accuracy for softer equations of 
state will take $15-20$ source detections; (ii) the statistical uncertainty
associated with $\lambdans$ measurement shrinks to within $\pm50\%$ with
$10-20$ events, and to within $\pm 25\%$ with $50+$ events, when source 
$\lambdans\geq 1000$; (iii) for softer equations of state, the same could take
$25-40$ and $50+$ events, respectively for the two uncertainty thresholds;
and (iv) if BHs really do observe the astrophysical mass-gap, the information
accumulation is somewhat slower than if they do not. We conclude that within
$20-30$ observations, aLIGO would begin to place very interesting bounds on 
the NS deformability, which would allow us to rule out or rank different
equations of state for neutron star matter.

@@

\prayush{[PK: To be continued after revision of the introduction.]}

@@

\prayush{
The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
%
\textbf{%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
}
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}.
}

@@

\red{Move to discussion}
\prayush{%
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
confidence level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.
}

>>>>

\red{SAY HERE THAT WE ONLY USE THE $l=|m|=2$ modes.}
\prayush{
This is possible because of our simplification of 
using only the dominant $l=|m|=2$ modes of the GW strain~\footnote{The 
$l=|m|=2$ multipoles of the GW strain, when decomposed in a spin $-2$ weighted
spherical harmonic basis, contain more than $99\%$ of the total signal power}
that makes various angle parameters completely degenerate with distance. 
}
% 
\subsection{Waveform Models}\label{s2:waveforms}

@@

% \textbf{Multiple identical sources at low SNR: }\label{s2:identical_multiple}

@@

While we discussed Eq.~\ref{eq:rmsSigmaIdenticalSources} for a population
of optimally oriented sources, it is valid more generally for a population
distributed uniformly in {\it effective} volume~\footnote{{\it effective} 
distance $D_\mathrm{eff}$ 
is a combination of distance to the source, its orientation, and its sky
location angles, and has a one-to-one correspondence with SNR for non-precessing
systems, for which these angles are constant in time. Effective volume is 
$\frac{4\pi}{3}D_\mathrm{eff}^3$.}~\cite{Markakis:2010mp}. However, it 
still only applies to sources with identical masses and spins, and we 
overcome this limitation by performing a fully Bayesian analysis next.

\textbf{Astrophysical source population: }\label{s2:astro_multiple}
Imagine that we have $N$ stretches of data, $d_1, d_2, \cdots, d_N$, each 
containing a single signal emitted by an NSBH binary. Each of these signals can
be characterized by the non-tidal source parameters
$\vec{\theta} := \{\mbh, \mns, \chibh, \chins, \vec{\alpha}\}$,
and $\{\lambdans\}$, where $\vec{\alpha}$ contains extrinsic parameters,
such as source distance, inclination, and sky location angles.
Further, let $K$ denote all of our collective prior knowledge, except for 
expectations on binary parameters themselves, which enter the following 
calculation explicitly. For instance, $K$ includes our assumption that all
NS's in a single population have the same deformability parameter $\lambdans$, 
and its cumulative measurement is therefore possible.
The probability distribution for $\lambdans$, given $N$ unique and
independent events, is
\begin{eqnarray}
 p(\lambdans |&& \hspace{-4mm}d_1, d_2, \cdots, d_N, K)\hspace{50mm}\nonumber\\ &=& \dfrac{p(d_1,d_2,\cdots,d_N |\lambdans , K)\,p(\lambdans|K)}{\int p(\lambdans |K) p(d_1,d_2,\cdots,d_N |\lambdans , K)\D\lambdans},\label{eq:p11}\\
  &=& \dfrac{p(\lambdans|K) \prod_{i} p(d_i|\lambdans, K)}{\int p(\lambdans ) p(d_1,d_2,\cdots,d_N |\lambdans , K)\D\lambdans},\label{eq:p12} \\
  &=& \dfrac{p(\lambdans|K) \prod_{i} \left( p(\lambdans |d_i, K)\dfrac{p(d_i)}{p(\lambdans|K)} \right)}{\int\, p(\lambdans|K )\, p(d_1,d_2,\cdots,d_N |\lambdans , K)\D\lambdans}\label{eq:p13};
\end{eqnarray}
where Eq.~\ref{eq:p11} and Eq.~\ref{eq:p13} are application of Bayes' theorem,
while Eq.~\ref{eq:p12} comes from the mutual independence of all events.
Assuming in addition that all events are {\it equally likely}: 
$p(d_i) = p(d_j) = p(d)$, we get
\begin{eqnarray}
 p(&&\hspace{-4mm}\lambdans | d_1, d_2, \cdots, d_N, K)\hspace{50mm}\nonumber\\
  &=& \dfrac{p(\lambdans) \left(\dfrac{p(d)}{p(\lambdans)}\right)^N\prod_{i} p(\lambdans |d_i, K) }{\int p(\lambdans ) p(d_1,d_2,\cdots,d_N |\lambdans , K)\D\lambdans}\label{eq:p21},\\
  &=& p(\lambdans)^{1-N}\times \dfrac{p(d)^N}{\int p(\lambdans) p(d_1, d_2, \cdots, d_N |\lambdans, K)\D\lambdans}\nonumber\\ &&\hspace{3mm}\times\prod_i p(\lambdans |d_i, K)\label{eq:p22}.
\end{eqnarray}
The first two factors in Eq.~\ref{eq:p22} can be absorbed into a
normalization factor $\mathcal{N}$, simplifying it to
\begin{equation}\label{eq:lambdaMultiple}
 p(\lambdans | d_1, d_2, \cdots, d_N; K) = \mathcal{N}\prod_{i=1}^N p(\lambdans | d_i, K).
% Using Bayes' theorem, the measured probability distribution function for 
% $\lambdans$ from $N$ observations is given by~\footnote{%
% Eq.~\ref{eq:lambdaMultiple} includes the following implicit factor to normalize
% its right-hand side: $\dfrac{\prod_i p(d_i|K)}{\int p(\lambdans)\,p(d_1, d_2, \cdots, d_N | \lambdans; K)\,\D\lambdans}$}
% \begin{equation}\label{eq:lambdaMultiple}
%  p(\lambdans | d_1, d_2, \cdots, d_N; K) = p(\lambdans | K)^{1-N}\prod_{i=1}^N p(\lambdans | d_i, K),
A priori, we assume that no particular value of $\lambdans$ is preferred
over another, and restrict $\lambdans$ to $[0, 4000]$, i.e.
\begin{equation}\label{eq:lprior}
 p(\lambdans | K) = \dfrac{1}{4000}\,\mathrm{Rect}\left(\frac{\lambdans-2000}{4000}\right).
In the second set of terms in Eq.~\ref{eq:lambdaMultiple} (of the form 
$p(\lambdans | d_i, K)$), each is the probability distribution for $\lambdans$
inferred {\it a posteriori} from the \textit{i}-th observation by marginalizing
\begin{equation}\label{eq:margpost}
 p(\lambdans | d_i, K) = \int\, p(\vec{\theta}, \lambdans | d_i, K)\, \D \vec{\theta},
where $p(\vec{\theta}, \lambdans | d_i, K)$ is the inferred joint probability 
distribution of all source parameters $\vec{\theta}\cup\{\lambdans\}$ for the 
$i$-th event, as given by Eq.~\ref{eq:postprob}. We note that 
Fig.~\ref{fig:SingleSystemLambdaPDFvsSNR} illustrates $p(\lambdans | d_i, K)$
for three individual events. By substituting
Eq.~\ref{eq:lprior}-\ref{eq:margpost} into Eq.~\ref{eq:lambdaMultiple}, we
calculate the probability distribution for $\lambdans$ as measured using $N$
independent events.

Our goal is to understand the improvement in our measurement of $\lambdans$
with the number of recorded events. To do so, we simulate a population~\footnote{%
A population here is an ordered set of events, and an event itself is the 
set of parameters describing one astrophysical NSBH binary.}
of $N$ events, and quantify what we learn from each successive observation 
using Eq.~\ref{eq:lambdaMultiple}. This allows us to quantify how
rapidly our median estimate for $\lambdans$ converges to the true value,
and how rapidly our confidence intervals for the same shrink, with increasing
$N$. Finally, we generate and analyze an ensemble of populations in order to
average over the stochastic process of population generation itself.

In order to generate each population, the first step is to fix
the NS properties: (i) NS mass $\mns=1.35M_\odot$, (ii) NS spin $\chins=0$
and (iii) NS tidal deformability $\lambdans=$ fixed value chosen from
$\{500,800,1000,1500,2000\}$. Next, we generate events, by sampling BH mass
(uniformly) from $\mbh\in[3M_\odot,6.75M_\odot]$, BH spin (uniformly)
from $\chibh\in[0, 0.75]$, source location uniformly in spatial volume,
and its inclination from $\iota\in[0,\pi]$. \prayush{Note that we only
sample positive aligned BH spins, since binaries with anti-aligned spins have
very little information to add at realistic SNRs, as shown in 
Fig.~\ref{fig:TT_LambdaCIWidths90_0_Lambda_SNR}.
}
As noted previously, we also restrict events to have
$\rho\geq\rho_\mathrm{min}$, where $\rho_\mathrm{min}=10$.
Finally, we repeat the process till we have an ordered set of $N$ events.
% 
Since we want to analyze not just a single realization of an astrophysical
population, but an ensemble of them, we make an additional approximation to
mitigate computational cost. Complete Bayesian parameter estimation is
performed for a set of simulated signals whose parameters are the vertices
of a regular hypercubic grid (henceforth ``G'') in the space of
$\{q\}\times\{\chibh\}\times\{\rho\}$, with each sampled at $q=\{2,3,4,5\}$,
$\chibh=\{-0.5,0,0.5,0.75\}$, and $\rho=\{10,20,30,50,70\}$.
All events in each population draw are substituted by their
respective nearest neighbours on the grid G.

In Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0} we show illustrative results
for a single population with neutron star deformability $\lambdans=800$.
In the left panel, each curve shows the 
probability distributions for $\lambdans$ as inferred from $N$ events, with $N$
ranging from $1-80$. We also mark the $90\%$ confidence intervals associated
with each of the probability distribution curves. The first few observations
do not have
enough information to bound $\lambdans$ much more than
our prior from Eq.~\ref{eq:lprior} does. However, the median of these
cumulative distributions appears to track the true value of $\lambdans$ 
rapidly.
% 
In the right panel, we present information derived from the left panel.
The line-circle curve shows the measured median from $N$ observations.
The pair of dashed (dotted) horizontal lines mark
$\pm25\%$ ($\pm50\%$) error bars. At each $N$, the range spanned by the 
filled region is the $90\%$ confidence interval deduced from the same 
events. This figure somewhat quantifies the qualitative deductions we made
from the left panel. We find that the measured median does track the true value
quickly, reaching $10\%$ accuracy with $10-15$ observations. With the same
information, our confidence intervals also shrink to $\pm 25\%$.
% 
In Fig.~\ref{fig:TT_Lambda_vs_N_CI90_0} we show further results from four
independent populations for $\lambdans=\{500,1000,1500,2000\}$. As in the 
right panel of Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}, the line-circle curves
track the median measured $\lambdans$, while the filled regions show
the associated $90\%$ confidence intervals. From the figure, we observe
the following: (i) the shrinkage of confidence interval widths with increasing
$N$ happens in a similar manner for each $\lambdans$, (ii) within $\sim 10$
observations, median measured $\lambdans$ for all populations lie within
$10\%$ of the true values, and (iii) it takes
approximately $20$ events to distinguish definitively (with $90\%$ confidence)
between deformable NSs with $\lambdans=2000$ and compact NSs with 
$\lambdans=500$, or equivalently to distinguish between hard, moderate and soft
nuclear equations of state. This is comparable to what has been found for
binary neutron stars~\cite{DelPozzo:13,Chatziioannou:2015uea,Agathos:2015a}.

So far we have discussed individual realizations of NSBH populations. The 
underlying stochasticity of the population generation process makes it
difficult to draw generalized inferences (from a single realization) about the
measurability of $\lambdans$. We mitigate this by drawing multiple
realizations of population generation (i.e. multiple identically constructed
populations). Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} shows the first
results for multiple populations. Lets focus on the left panel first. We
show the median measured value of $\lambdans$ from $100$ independent 
populations with fixed $\lambdans=2000$ as the top set of curves. The one 
highlighted (in green) curve corresponds to the population discussed above.
Similarly, we generate sets of $100$ populations with $\lambdans=500-1500$, and
show the median measured $\lambdans$ as four distinct groups of curves in
Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} (with the single highlighted
curve in each group corresponding to
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}).
Dash-dotted horizontal lines demarcate $\pm50\%$ error intervals around
the true $\lambdans$ values. All of these populations are drawn under
paradigm A (i.e. with BH masses allowed to sample the mass-gap that spans
$2-5M_\odot$).
% 
From the figure, we can conclude with a high likelihood that our measured
$\lambdans$ values will be within $10\%$ of the real value after about
$20$ events for less deformable neutron stars with $\lambdans\leq 1000$,
and in as few as $10$ events for more deformable stars with
$\lambdans\geq 1500$.
% 
The results discussed in Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0},
\ref{fig:TT_Lambda_vs_N_CI90_0} and the left panel
of Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} all apply to the parameter distribution spanned by 
the grid G. This distribution has $\mbh$ as low as $2.7M_\odot$ (i.e. $q=2$).
Given that disruptive signatures are strongest for small $\mbh$, we now
investigate an alternate paradigm~\footnote{We denote our standard paradigm
which does not respect the mass-gap as paradigm A; with paradigm B being
the scenario where no black hole masses fall within the mass gap.}
with $\mbh\geq 5M_\odot$, consistent
with the ``mass-gap'' suggested by astronomical observations~\cite{Bailyn:1997xt,
Kalogera:1996ci,Kreidberg:2012,Littenberg:2015tpa}
% 
The right panel of the figure is identical to the left panel, but drawn under
population paradigm B (no BHs alloewd within the mass-gap). Under this
paradigm, we expectedly find that information accumulation is slower. It would
take $20-40$ detections with $\rho\geq10$ under this paradigm, for our (median)
measured $\lambdans$ to lie within $10\%$ of the true value (fewer for 
more deformable NSs).
% % 
% This is a promising deduction, as one might expect to see $\mathcal{O}(10)$
% events over a few years' timescale with design aLIGO~\cite{Abadie:2010cfa}.

Finally, we investigate the statistical uncertainties associated with
$\lambdans$ measurements. We use $90\%$ confidence intervals as our measure of
the same. The left column of 
Fig.~\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne} shows results
under paradigm A. In each panel
we show the width of the $90\%$ confidence interval measured from $N$ events
(with $N$ on the $x$-axis), normalized by the true value of $\lambdans$ for
the population. In gray are shown results from each of $100$ population
realizations, with the single population highlighted in color corresponding
to the one we focused on in 
Fig.~\ref{fig:TT_Lambda_vs_N_L800_CI90_0}-\ref{fig:TT_Lambda_vs_N_CI90_0}.
The inset shows the same, except that the ordinate is {\it not} normalized by
the true value of $\lambdans$. From top to bottom, the population $\lambdans$
decreases from $\lambdans=2000\rightarrow 500$, corresponding to decreasingly
deformable NSs with softer equations of state. We observe the following:
(i) for moderately-hard to hard equations of state with $\lambdans\geq 1000$,
we can constrain $\lambdans$ within $\pm 50\%$ using only $10-20$ events, and
within $\pm 25\%$ with $25-40$ events; (ii) for softer equations of state with
$\lambdans<1000$, we will achieve the same accuracy with $20-30$ and $50+$ 
events, respectively; and (iii) for the first $5$ or so observations, our
measurement confidence spans the entire prior allowed range 
$\lambdans\in[0,4000]$, as shown by the plateauing towards the left edge in
each panel's inset.
% 
In the right column, each panel panel is identical to its corresponding left
one, with the difference that populations are drawn under paradigm B, which
does {\it not} allow for BH masses to fall within the mass-gap. We find that
for NSs with $\lambdans\leq 1000$, it would take $25-40$ events to
constrain $\lambdans$ within $\pm 50\%$ and $50+$ events to constrain it
within $\pm 25\%$. This is somewhat slower than paradigm A, as is to be
expected since here we preclude the lowest mass-ratios, which correspond to
signals with largest tidal signatures. For $\lambdans>1000$ we find that we
can constrain $\lambdans$ within $\pm 50\%$ with a similar number of events as
for paradigm A, but will need more ($30-40$, as compared to $25-40$) events
to further constrain it to within $\pm 25\%$ of the true value.
%

@@

To summarize, in this section we study the improvement in our measurement of 
NS deformability parameter $\lambdans$ with an increasing number of events. We
do so by simulating plausible populations of disrupting NSBH binaries (with
$\rho\geq 10$). We find that:
(i) for more deformable neutron stars (harder equation of states), the median
measured value of $\lambdans$ comes within $10\%$ of the true value with as 
few as $10$ events, while achieving the same accuracy for softer equations of 
state will take $15-20$ source detections; (ii) the statistical uncertainty
associated with $\lambdans$ measurement shrinks to within $\pm50\%$ with
$10-20$ events, and to within $\pm 25\%$ with $50+$ events, when source 
$\lambdans\geq 1000$; (iii) for softer equations of state, the same could take
$25-40$ and $50+$ events, respectively for the two uncertainty thresholds;
and (iv) if BHs really do observe the astrophysical mass-gap, the information
accumulation is somewhat slower than if they do not. We conclude that within
$20-30$ observations, aLIGO would begin to place very interesting bounds on 
the NS deformability, which would allow us to rule out or rank different
equations of state for neutron star matter. Our key findings are 
summarized in Fig.~\ref{fig:TT_LambdaMedian_vs_N_AllInOne} and
\ref{fig:TT_LambdaError_vs_N_L500_2000_CI90_0_AllInOne}.

@@

@@

\prayush{
The choices that determine the population are important here,
and were made as follows. Neutron star mass is fixed for signals at
$1.35M_\odot$, and its spin $\chins=0$. Black hole mass is sampled uniformly
from the range $[2,5]\times 1.35=[2.7, 6.75]M_\odot$. Our choice here is given
by the intersection set of the mass range that allows for neutron star disruption
and the range supported by the waveform model LEA+~\cite{Foucart2012,
Foucart:2013a,Lackey:2013axa}. Black hole spin is sampled uniformly from the
{\it aligned} range, i.e. $\chibh\in[0,+1]$. The true deformability is fixed
once per population, sampled uniformly from the range $\lambdans\in[0,2000]$.
In order to keep the computational cost tractable, we make a further 
approximation.
%
We inject signals into zero noise at different SNRs and recover
their parameters from Bayesian inferencing for a fixed set of binaries,
whose parameters lie on a uniform grid given by:
$q=\{2,3,4,5\}\times\chibh=\{-0.5,0,0.5,0.75\}\times\lambdans=\{500,800,1000,1500,2000\}\times\rho=\{10,20,30,50,70\}$.
% 
Once the population parameters been selected, each event's parameters are
replaced by their nearest neighbor in the above set.
% 
We make sure that this approximation is conservative by keeping our population
sampling ranges within the conservative side of the range spanned by the above
set. With the caveats laid out, we report our results in
Sec.~\ref{s1:multiple_observations}.
}

@@

\prayush{%
We find that when the BH is sufficiently spinning in the prograde sense,
with $\chibh\gtrsim +0.7$, even at a relatively possible signal strengths ($\rho=20$),
we can begin to constraint $\lambdans$ at the level of $\pm 75\%$ errors with a $90\%$
confidence level. If the signal is stronger, say $\rho=30$, we can contraint
$\lambdans$ better with only $\pm 50\%$ errors, for binaries with large BH spins.
This trend continues as we vary the SNR from $\rho=30-50$ upwards. In other words,
with a single but moderately loud NSBH signal observation, Advanced LIGO will begin
to put constraints on NS matter properties. These constraints can consequently
be used to assess the likelihood of different equations of state for nuclear
matter, and possibly narrow the range they span.
}
